---
title: "RIPTA Data Merge - Main Pipeline"
author: "Zhiqian He"
date: "2026-03-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r libraries}
library(data.table)
library(sf)
library(ggplot2)
```

# Configuration

```{r config}
# =============================
# Main paths
# =============================
base_data_dir <- "/Users/erica/Desktop/RIPTA/data"
base_code_dir <- "/Users/erica/Desktop/RIPTA/code/2026.03.10 Merge monthly data"

ridership_file <- file.path(base_data_dir, "Ridership Data", "april24_preprocessed.csv")
system_file    <- file.path(base_code_dir, "AVLOTP2024.csv")

# GTFS version parameter
GTFS_TAG <- "20240112"
stops_file <- file.path(base_code_dir, paste0("stops_RIPTA-GTFS_", GTFS_TAG, ".txt"))

# Analysis month parameter
ANALYSIS_MONTH <- "2024-04"
ANALYSIS_TAG   <- "april2024"

# Output folder / names
output_dir <- file.path(base_code_dir, paste0("outputs_", ANALYSIS_TAG, "_gtfs", GTFS_TAG))
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

mapped_avl_file     <- file.path(output_dir, paste0("mapped_avl_", ANALYSIS_TAG, "_gtfs", GTFS_TAG, ".csv"))
daily_match_file    <- file.path(output_dir, paste0("daily_match_", ANALYSIS_TAG, "_gtfs", GTFS_TAG, ".csv"))
monthly_bridge_file <- file.path(output_dir, paste0("monthly_bridge_", ANALYSIS_TAG, "_gtfs", GTFS_TAG, ".csv"))
route_summary_file  <- file.path(output_dir, paste0("route_otp_summary_", ANALYSIS_TAG, "_gtfs", GTFS_TAG, ".csv"))
scatter_plot_file   <- file.path(output_dir, paste0("otp_vs_weighted_otp_", ANALYSIS_TAG, "_gtfs", GTFS_TAG, ".png"))
trip_rankings_file  <- file.path(output_dir, paste0("trip_rankings_", ANALYSIS_TAG, "_gtfs", GTFS_TAG, ".csv"))

# Parameters
IOU_MIN <- 0.10
TOLERANCE_SEC <- 120
MAX_HEADWAY_SEC <- 1800
KEEP_MAPPING_FLAGS <- c("good", "ok")
BAD_APP_IDS <- c(0L, 99L, 999L, 9999L)
OTP_THRESHOLD_MIN <- 5
MIN_DAYS_FOR_RANKING <- 3

# Helpers
seconds_of_day <- function(dt) {
  as.integer(format(dt, "%H")) * 3600L +
    as.integer(format(dt, "%M")) * 60L +
    as.integer(format(dt, "%S"))
}

format_seconds <- function(x) {
  x_int <- as.integer(round(x))
  ifelse(
    is.na(x_int), "",
    sprintf("%02d:%02d:%02d",
            x_int %/% 3600L,
            (x_int %% 3600L) %/% 60L,
            x_int %% 60L)
  )
}

calculate_iou_containment <- function(start1, end1, start2, end2) {
  overlap_start <- pmax(start1, start2)
  overlap_end   <- pmin(end1, end2)
  intersection  <- pmax(0, overlap_end - overlap_start)
  duration1 <- pmax(0, end1 - start1)
  duration2 <- pmax(0, end2 - start2)
  union <- duration1 + duration2 - intersection
  iou <- ifelse(union > 0, intersection / union, 0)
  smaller_duration <- pmin(duration1, duration2)
  containment <- ifelse(smaller_duration > 0, intersection / smaller_duration, 0)
  list(iou = iou, containment = containment)
}

determine_confidence <- function(ratio_within_2min, iou, containment,
                                 thresh_highly_conf = 0.8,
                                 thresh_conf = 0.5,
                                 thresh_contain = 0.9) {
  if (!is.na(ratio_within_2min) && ratio_within_2min >= 0.5) {
    "Highly Confident (Stop-Validated)"
  } else if (!is.na(iou) && iou >= thresh_highly_conf) {
    "Highly Confident"
  } else if (!is.na(iou) && !is.na(containment) &&
             iou >= thresh_conf && containment >= thresh_contain) {
    "Confident"
  } else if (!is.na(iou) && iou >= thresh_conf) {
    "Confident"
  } else {
    "Uncertain"
  }
}
```

# Step 1. Map AVL stops to GTFS stops

```{r step1-map-avl}
rid <- fread(file = ridership_file)
sys <- fread(file = system_file)
stops <- fread(file = stops_file)

rid[, ServiceDate := as.Date(Time)]
sys[, ServiceDate := as.Date(IncidentDateTime)]
sys_apr <- sys[format(ServiceDate, "%Y-%m") == ANALYSIS_MONTH]

stops_small <- stops[
  !is.na(stop_lat) & !is.na(stop_lon),
  c("stop_id", "stop_code", "stop_name", "stop_lat", "stop_lon"),
  with = FALSE
]

sys_pts <- unique(
  sys_apr[
    !is.na(BusStopLatitude) & !is.na(BusStopLongitude),
    c("BusStopLatitude", "BusStopLongitude"),
    with = FALSE
  ]
)

sys_sf <- st_as_sf(sys_pts, coords = c("BusStopLongitude", "BusStopLatitude"), crs = 4326, remove = FALSE)
stops_sf <- st_as_sf(stops_small, coords = c("stop_lon", "stop_lat"), crs = 4326, remove = FALSE)

sys_sf_m <- st_transform(sys_sf, 3857)
stops_sf_m <- st_transform(stops_sf, 3857)
nearest_idx <- st_nearest_feature(sys_sf_m, stops_sf_m)
nearest_stop <- stops_small[nearest_idx, ]
dist_m <- as.numeric(st_distance(sys_sf_m, stops_sf_m[nearest_idx, ], by_element = TRUE))

sys_map <- cbind(as.data.table(sys_pts), as.data.table(nearest_stop), dist_m = dist_m)
sys_map[, mapping_flag := fifelse(
  is.na(dist_m), "missing_coord",
  fifelse(dist_m <= 50, "good",
          fifelse(dist_m <= 200, "ok", "far"))
)]

sys_apr_mapped <- merge(
  sys_apr,
  sys_map,
  by = c("BusStopLatitude", "BusStopLongitude"),
  all.x = TRUE
)
setnames(sys_apr_mapped,
         old = c("stop_id", "stop_code"),
         new = c("Mapped_StopId", "Mapped_StopCode"),
         skip_absent = TRUE)
sys_apr_mapped[is.na(BusStopLatitude) | is.na(BusStopLongitude), mapping_flag := "missing_coord"]

fwrite(sys_apr_mapped, file = mapped_avl_file)

cat("Saved mapped AVL file:\n", mapped_avl_file, "\n")
print(sys_apr_mapped[, .N, by = mapping_flag][order(mapping_flag)])

rm(rid, sys, stops, sys_apr, stops_small, sys_pts,
   sys_sf, stops_sf, sys_sf_m, stops_sf_m,
   nearest_idx, nearest_stop, dist_m, sys_map)
gc()
```

# Step 2. Daily matching between ridership and AVL

```{r step2-daily-match}
rid_raw <- fread(file = ridership_file)
sys_raw <- fread(file = mapped_avl_file)

setnames(
  rid_raw,
  old = c("Trip.Number", "Stop.Number", "Route.Number", "Ride.Count", "Time"),
  new = c("AppID", "app_stopID", "Route", "RideCount", "RideTime"),
  skip_absent = TRUE
)
if (!"ServiceDate" %in% names(rid_raw)) rid_raw[, ServiceDate := as.Date(RideTime)]
rid_raw[, RideTime := as.POSIXct(RideTime, tz = "America/New_York")]
rid_raw[, Route := as.integer(Route)]
rid_raw[, AppID := as.integer(AppID)]
rid_raw[, app_stopID := as.integer(app_stopID)]
rid_raw[, RideCount := as.numeric(RideCount)]
rid_raw[, Seconds := seconds_of_day(RideTime)]

rid_clean <- rid_raw[
  format(ServiceDate, "%Y-%m") == ANALYSIS_MONTH &
    !is.na(AppID) &
    !(AppID %in% BAD_APP_IDS),
  c("ServiceDate", "Route", "AppID", "app_stopID", "RideTime", "Seconds", "RideCount"),
  with = FALSE
]

setnames(
  sys_raw,
  old = c("RouteId", "IncidentDateTime", "Deviation"),
  new = c("Route", "SystemTime", "Deviation"),
  skip_absent = TRUE
)
if (!"ServiceDate" %in% names(sys_raw)) sys_raw[, ServiceDate := as.Date(SystemTime)]
sys_raw[, SystemTime := as.POSIXct(SystemTime, tz = "America/New_York")]
sys_raw[, Route := as.integer(Route)]
sys_raw[, SystemID := as.integer(SystemID)]
sys_raw[, Mapped_StopId := as.integer(Mapped_StopId)]
sys_raw[, Deviation := as.numeric(Deviation)]
sys_raw[, Seconds := seconds_of_day(SystemTime)]

sys_clean <- sys_raw[
  format(ServiceDate, "%Y-%m") == ANALYSIS_MONTH &
    mapping_flag %in% KEEP_MAPPING_FLAGS &
    !is.na(Mapped_StopId),
  c("ServiceDate", "Route", "SystemID", "Mapped_StopId", "SystemTime",
    "Seconds", "Deviation", "mapping_flag", "dist_m", "IncidentName"),
  with = FALSE
]

app_profile_daily <- rid_clean[
  , .(
    App_Start_Sec = min(Seconds, na.rm = TRUE),
    App_End_Sec   = max(Seconds, na.rm = TRUE),
    Number_of_Riders = sum(RideCount, na.rm = TRUE)
  ),
  by = .(Route, ServiceDate, AppID)
]

app_stop_profile_daily <- rid_clean[
  , .(App_StopTime_Median = median(Seconds, na.rm = TRUE)),
  by = .(Route, ServiceDate, AppID, app_stopID)
]

sys_profile_daily <- sys_clean[
  , .(
    Sys_Start_Sec = min(Seconds, na.rm = TRUE),
    Sys_End_Sec   = max(Seconds, na.rm = TRUE)
  ),
  by = .(Route, ServiceDate, SystemID)
]

sys_stop_profile_daily <- sys_clean[
  , .(Sys_StopTime_Median = median(Seconds, na.rm = TRUE)),
  by = .(Route, ServiceDate, SystemID, Mapped_StopId)
]

match_one_route_day <- function(route_val, date_val,
                                app_profile_daily,
                                sys_profile_daily,
                                app_stop_profile_daily,
                                sys_stop_profile_daily) {
  app_r <- app_profile_daily[Route == route_val & ServiceDate == date_val]
  sys_r <- sys_profile_daily[Route == route_val & ServiceDate == date_val]
  if (nrow(app_r) == 0 || nrow(sys_r) == 0) return(NULL)

  app_r[, join_key := 1L]
  sys_r[, join_key := 1L]
  cand <- merge(app_r, sys_r, by = "join_key", allow.cartesian = TRUE)
  cand[, join_key := NULL]
  if (nrow(cand) == 0) return(NULL)

  if ("Route.x" %in% names(cand)) setnames(cand, "Route.x", "Route")
  if ("Route.y" %in% names(cand)) cand[, Route.y := NULL]
  if ("ServiceDate.x" %in% names(cand)) setnames(cand, "ServiceDate.x", "ServiceDate")
  if ("ServiceDate.y" %in% names(cand)) cand[, ServiceDate.y := NULL]

  m <- calculate_iou_containment(cand$Sys_Start_Sec, cand$Sys_End_Sec,
                                 cand$App_Start_Sec, cand$App_End_Sec)
  cand[, IoU := m$iou]
  cand[, Containment := m$containment]
  cand <- cand[IoU > IOU_MIN]
  if (nrow(cand) == 0) return(NULL)

  app_stops_r <- app_stop_profile_daily[Route == route_val & ServiceDate == date_val]
  sys_stops_r <- sys_stop_profile_daily[Route == route_val & ServiceDate == date_val]

  stop_eval <- NULL
  if (nrow(app_stops_r) > 0 && nrow(sys_stops_r) > 0) {
    merged_stops <- merge(
      app_stops_r,
      sys_stops_r,
      by.x = c("Route", "ServiceDate", "app_stopID"),
      by.y = c("Route", "ServiceDate", "Mapped_StopId"),
      allow.cartesian = TRUE
    )

    if (nrow(merged_stops) > 0) {
      candidate_keys <- unique(cand[, c("Route", "ServiceDate", "AppID", "SystemID"), with = FALSE])
      merged_stops <- merge(merged_stops, candidate_keys,
                            by = c("Route", "ServiceDate", "AppID", "SystemID"))
      if (nrow(merged_stops) > 0) {
        merged_stops[, time_diff := abs(App_StopTime_Median - Sys_StopTime_Median)]
        stop_eval <- merged_stops[
          , .(
            Max_Stop_Diff = max(time_diff, na.rm = TRUE),
            Ratio_Within_2Min = mean(time_diff <= TOLERANCE_SEC, na.rm = TRUE),
            Common_Stops = .N
          ),
          by = .(Route, ServiceDate, AppID, SystemID)
        ]
      }
    }
  }

  if (is.null(stop_eval) || nrow(stop_eval) == 0) {
    cand[, Max_Stop_Diff := NA_real_]
    cand[, Ratio_Within_2Min := 0]
    cand[, Common_Stops := 0L]
  } else {
    cand <- merge(cand, stop_eval,
                  by = c("Route", "ServiceDate", "AppID", "SystemID"), all.x = TRUE)
    cand[is.na(Max_Stop_Diff), Max_Stop_Diff := NA_real_]
    cand[is.na(Ratio_Within_2Min), Ratio_Within_2Min := 0]
    cand[is.na(Common_Stops), Common_Stops := 0L]
  }

  valid_candidates <- cand[
    !is.na(Max_Stop_Diff) & Max_Stop_Diff <= MAX_HEADWAY_SEC & Common_Stops > 0
  ]

  if (nrow(valid_candidates) == 0) {
    setorder(cand, AppID, -IoU)
    best_matches <- cand[, .SD[1], by = AppID]
  } else {
    setorder(valid_candidates, AppID, -Ratio_Within_2Min, -IoU)
    best_matches <- valid_candidates[, .SD[1], by = AppID]
  }

  best_matches[, Level_of_Confidence :=
                 mapply(determine_confidence, Ratio_Within_2Min, IoU, Containment)]
  best_matches[, App_Start_Time := format_seconds(App_Start_Sec)]
  best_matches[, App_End_Time   := format_seconds(App_End_Sec)]
  best_matches[, Sys_Start_Time := format_seconds(Sys_Start_Sec)]
  best_matches[, Sys_End_Time   := format_seconds(Sys_End_Sec)]
  best_matches[]
}

route_day_keys <- unique(app_profile_daily[, c("Route", "ServiceDate"), with = FALSE])
route_day_keys <- merge(route_day_keys,
                        unique(sys_profile_daily[, c("Route", "ServiceDate"), with = FALSE]),
                        by = c("Route", "ServiceDate"))
setorder(route_day_keys, Route, ServiceDate)

all_matches_list <- vector("list", nrow(route_day_keys))
for (i in seq_len(nrow(route_day_keys))) {
  all_matches_list[[i]] <- match_one_route_day(
    route_val = route_day_keys$Route[i],
    date_val = route_day_keys$ServiceDate[i],
    app_profile_daily = app_profile_daily,
    sys_profile_daily = sys_profile_daily,
    app_stop_profile_daily = app_stop_profile_daily,
    sys_stop_profile_daily = sys_stop_profile_daily
  )
}

matched_trips_daily <- rbindlist(all_matches_list, use.names = TRUE, fill = TRUE)
matched_trips_daily <- matched_trips_daily[, .(
  Route, ServiceDate, AppID, Number_of_Riders, SystemID,
  Level_of_Confidence, IoU, Containment,
  Ratio_Within_2Min, Max_Stop_Diff, Common_Stops,
  App_Start_Time, App_End_Time, Sys_Start_Time, Sys_End_Time,
  App_Start_Sec, App_End_Sec, Sys_Start_Sec, Sys_End_Sec
)]
setorder(matched_trips_daily, Route, ServiceDate, AppID)

fwrite(matched_trips_daily, file = daily_match_file)
cat("Saved daily match file:\n", daily_match_file, "\n")
cat("Matched trip occurrences:", nrow(matched_trips_daily), "\n")

rm(rid_raw, sys_raw, rid_clean, sys_clean,
   app_profile_daily, app_stop_profile_daily,
   sys_profile_daily, sys_stop_profile_daily,
   route_day_keys, all_matches_list)
gc()
```

# Step 2.2. Monthly 1-to-1 bridge

```{r step22-monthly-bridge}
match_daily <- fread(file = daily_match_file)

match_daily[, confidence_score := fifelse(
  Level_of_Confidence == "Highly Confident (Stop-Validated)", 4,
  fifelse(Level_of_Confidence == "Highly Confident", 3,
          fifelse(Level_of_Confidence == "Confident", 2, 1))
)]

pair_summary <- match_daily[
  , .(
    matched_days = .N,
    total_riders_month = sum(Number_of_Riders, na.rm = TRUE),
    avg_riders_per_matched_day = mean(Number_of_Riders, na.rm = TRUE),
    mean_iou = mean(IoU, na.rm = TRUE),
    mean_containment = mean(Containment, na.rm = TRUE),
    mean_ratio_2min = mean(Ratio_Within_2Min, na.rm = TRUE),
    mean_common_stops = mean(Common_Stops, na.rm = TRUE),
    mean_confidence_score = mean(confidence_score, na.rm = TRUE)
  ),
  by = .(Route, AppID, SystemID)
]

app_summary <- match_daily[
  , .(
    app_days_observed = uniqueN(ServiceDate),
    total_riders_month_app = sum(Number_of_Riders, na.rm = TRUE)
  ),
  by = .(Route, AppID)
]

system_variety <- pair_summary[
  , .(unique_system_ids_seen = uniqueN(SystemID)),
  by = .(Route, AppID)
]

setorder(pair_summary, Route, AppID,
         -matched_days, -mean_confidence_score, -mean_iou, -mean_ratio_2min)
monthly_bridge <- pair_summary[, .SD[1], by = .(Route, AppID)]
monthly_bridge <- merge(monthly_bridge, app_summary, by = c("Route", "AppID"), all.x = TRUE)
monthly_bridge <- merge(monthly_bridge, system_variety, by = c("Route", "AppID"), all.x = TRUE)
monthly_bridge[, dominant_share := matched_days / app_days_observed]
monthly_bridge[, Monthly_Match_Stability := fifelse(
  dominant_share >= 0.8 & unique_system_ids_seen == 1, "Very Stable",
  fifelse(dominant_share >= 0.6, "Mostly Stable",
          fifelse(dominant_share >= 0.4, "Mixed", "Unstable"))
)]

monthly_bridge <- monthly_bridge[, .(
  Route, AppID, SystemID,
  app_days_observed, matched_days, dominant_share,
  unique_system_ids_seen, Monthly_Match_Stability,
  total_riders_month, avg_riders_per_matched_day,
  mean_iou, mean_containment, mean_ratio_2min,
  mean_common_stops, mean_confidence_score,
  total_riders_month_app
)]
setorder(monthly_bridge, Route, AppID)

fwrite(monthly_bridge, file = monthly_bridge_file)
cat("Saved monthly bridge file:\n", monthly_bridge_file, "\n")

rm(match_daily, pair_summary, app_summary, system_variety)
gc()
```

# Step 3. OTP summary and trip rankings

```{r step3-otp-and-rankings}
matched_daily <- fread(file = daily_match_file)
sys_for_otp <- fread(file = mapped_avl_file)
monthly_bridge <- fread(file = monthly_bridge_file)

matched_daily[, ServiceDate := as.Date(ServiceDate)]
sys_for_otp[, ServiceDate := as.Date(ServiceDate)]

setnames(
  sys_for_otp,
  old = c("RouteId", "IncidentDateTime", "TripId", "Deviation"),
  new = c("Route", "SystemTime", "SystemID", "Deviation"),
  skip_absent = TRUE
)
sys_for_otp[, Route := as.integer(Route)]
sys_for_otp[, SystemID := as.integer(SystemID)]
sys_for_otp[, Mapped_StopId := as.integer(Mapped_StopId)]
sys_for_otp[, Deviation := as.numeric(Deviation)]

sys_for_otp <- sys_for_otp[
  format(ServiceDate, "%Y-%m") == ANALYSIS_MONTH &
    mapping_flag %in% KEEP_MAPPING_FLAGS &
    !is.na(Mapped_StopId)
]

# Team-level weighted OTP definition used for final results:
# trip_on_time is binary at trip-occurrence level, then weighted by Number_of_Riders.
sys_for_otp[, on_time_point := as.integer(abs(Deviation) <= OTP_THRESHOLD_MIN)]
trip_occurrence <- sys_for_otp[
  , .(
    trip_otp = as.numeric(mean(on_time_point, na.rm = TRUE)),
    mean_abs_deviation = as.numeric(mean(abs(Deviation), na.rm = TRUE)),
    n_timepoints = as.integer(.N)
  ),
  by = .(Route, ServiceDate, SystemID)
]
trip_occurrence[, trip_on_time := as.integer(trip_otp >= 0.5)]

matched_otp <- merge(
  matched_daily,
  trip_occurrence,
  by = c("Route", "ServiceDate", "SystemID"),
  all.x = TRUE
)
matched_otp <- matched_otp[!is.na(trip_on_time)]

route_summary <- matched_otp[
  , .(
    matched_trip_occurrences = as.integer(.N),
    total_riders = as.numeric(sum(Number_of_Riders, na.rm = TRUE)),
    avg_riders_per_occurrence = as.numeric(mean(Number_of_Riders, na.rm = TRUE)),
    OTP = as.numeric(mean(trip_on_time, na.rm = TRUE)),
    Weighted_OTP = as.numeric(
      sum(Number_of_Riders * trip_on_time, na.rm = TRUE) /
        sum(Number_of_Riders, na.rm = TRUE)
    ),
    mean_trip_otp = as.numeric(mean(trip_otp, na.rm = TRUE)),
    weighted_mean_trip_otp = as.numeric(
      sum(Number_of_Riders * trip_otp, na.rm = TRUE) /
        sum(Number_of_Riders, na.rm = TRUE)
    )
  ),
  by = Route
]
route_summary[, OTP_pct := 100 * OTP]
route_summary[, Weighted_OTP_pct := 100 * Weighted_OTP]
route_summary[, mean_trip_otp_pct := 100 * mean_trip_otp]
route_summary[, weighted_mean_trip_otp_pct := 100 * weighted_mean_trip_otp]
setorder(route_summary, Route)
fwrite(route_summary, file = route_summary_file)

p <- ggplot(route_summary, aes(x = OTP, y = Weighted_OTP, size = total_riders, label = Route)) +
  geom_point(alpha = 0.7) +
  geom_text(nudge_y = 0.01, size = 3) +
  labs(
    title = paste0("OTP vs Weighted OTP by Route (", ANALYSIS_MONTH, ")"),
    x = "OTP",
    y = "Weighted OTP",
    size = "Total Riders"
  ) +
  theme_minimal()

ggsave(filename = scatter_plot_file, plot = p, width = 8, height = 6, dpi = 300)

trip_daily <- merge(
  matched_daily,
  trip_occurrence,
  by = c("Route", "ServiceDate", "SystemID"),
  all.x = TRUE
)
trip_daily <- trip_daily[!is.na(trip_on_time)]

trip_summary <- trip_daily[
  , .(
    matched_days = as.integer(uniqueN(ServiceDate)),
    total_riders_month = as.numeric(sum(Number_of_Riders, na.rm = TRUE)),
    avg_riders_per_day = as.numeric(mean(Number_of_Riders, na.rm = TRUE)),
    late_day_rate = as.numeric(mean(trip_on_time == 0, na.rm = TRUE)),
    on_time_day_rate = as.numeric(mean(trip_on_time == 1, na.rm = TRUE)),
    mean_trip_otp = as.numeric(mean(trip_otp, na.rm = TRUE)),
    mean_abs_deviation = as.numeric(mean(mean_abs_deviation, na.rm = TRUE))
  ),
  by = .(Route, AppID)
]

monthly_bridge_keep <- monthly_bridge[, .(
  Route, AppID, SystemID, Monthly_Match_Stability, dominant_share
)]
trip_summary <- merge(trip_summary, monthly_bridge_keep,
                      by = c("Route", "AppID"), all.x = TRUE)
trip_summary[, late_day_rate_pct := 100 * late_day_rate]
trip_summary[, mean_trip_otp_pct := 100 * mean_trip_otp]

trip_summary_rankable <- trip_summary[matched_days >= MIN_DAYS_FOR_RANKING]

top_late <- trip_summary_rankable[
  order(-late_day_rate, mean_trip_otp, -matched_days, -total_riders_month)
][1:min(20, .N)]
top_late[, ranking_type := "Most Often Late"]

top_ridership <- trip_summary[
  order(-total_riders_month, -avg_riders_per_day, -matched_days)
][1:min(20, .N)]
top_ridership[, ranking_type := "Highest Ridership"]

top_reliability <- trip_summary_rankable[
  order(mean_trip_otp, -late_day_rate, -matched_days, -total_riders_month)
][1:min(20, .N)]
top_reliability[, ranking_type := "Worst Reliability"]

trip_rankings <- rbindlist(
  list(top_late, top_ridership, top_reliability),
  use.names = TRUE,
  fill = TRUE
)
trip_rankings <- trip_rankings[, .(
  ranking_type, Route, AppID, SystemID, matched_days,
  total_riders_month, avg_riders_per_day,
  late_day_rate_pct, mean_trip_otp_pct,
  Monthly_Match_Stability, dominant_share
)]

fwrite(trip_rankings, file = trip_rankings_file)

cat("Saved route summary file:\n", route_summary_file, "\n")
cat("Saved scatter plot:\n", scatter_plot_file, "\n")
cat("Saved trip rankings file:\n", trip_rankings_file, "\n")
print(route_summary[1:10])
print(trip_rankings[, .N, by = ranking_type])

rm(sys_for_otp, trip_occurrence, matched_otp, p,
   trip_daily, trip_summary, trip_summary_rankable,
   top_late, top_ridership, top_reliability,
   monthly_bridge_keep)
gc()
```


# Step 4. Summary and Visualization

```{r}
# =========================================================
# Monthly bridge based visualization
# x = monthly trip-based OTP
# y = passenger-weighted monthly OTP
# =========================================================

monthly_bridge <- fread(file = monthly_bridge_file)
sys_for_monthly <- fread(file = mapped_avl_file)

monthly_bridge[, Route := as.integer(Route)]
monthly_bridge[, AppID := as.integer(AppID)]
monthly_bridge[, SystemID := as.integer(SystemID)]

# Rebuild clean AVL subset
setnames(
  sys_for_monthly,
  old = c("RouteId", "IncidentDateTime", "TripId", "Deviation"),
  new = c("Route", "SystemTime", "SystemID", "Deviation"),
  skip_absent = TRUE
)

sys_for_monthly[, ServiceDate := as.Date(ServiceDate)]
sys_for_monthly[, Route := as.integer(Route)]
sys_for_monthly[, SystemID := as.integer(SystemID)]
sys_for_monthly[, Mapped_StopId := as.integer(Mapped_StopId)]
sys_for_monthly[, Deviation := as.numeric(Deviation)]

sys_for_monthly <- sys_for_monthly[
  format(ServiceDate, "%Y-%m") == "2024-04" &
    mapping_flag %in% KEEP_MAPPING_FLAGS &
    !is.na(Mapped_StopId)
]

# Point-level on-time
sys_for_monthly[, on_time_point := as.integer(abs(Deviation) <= OTP_THRESHOLD_MIN)]

# Monthly system trip OTP
# one row = Route + SystemID across April
system_monthly_otp <- sys_for_monthly[
  ,
  .(
    monthly_trip_otp = as.numeric(mean(on_time_point, na.rm = TRUE)),
    mean_abs_deviation = as.numeric(mean(abs(Deviation), na.rm = TRUE)),
    n_points = as.integer(.N),
    observed_days = as.integer(uniqueN(ServiceDate))
  ),
  by = .(Route, SystemID)
]

# Optional: keep only more stable monthly bridges
# You can relax this if needed
monthly_bridge_use <- monthly_bridge[
  !is.na(SystemID) &
    Monthly_Match_Stability %in% c("Very Stable", "Mostly Stable")
]

# Join monthly bridge with monthly OTP
monthly_trip_join <- merge(
  monthly_bridge_use,
  system_monthly_otp,
  by = c("Route", "SystemID"),
  all.x = TRUE
)

monthly_trip_join <- monthly_trip_join[!is.na(monthly_trip_otp)]

# Route-level monthly summary
route_monthly_summary <- monthly_trip_join[
  ,
  .(
    n_trip_patterns = as.integer(.N),
    total_riders_month = as.numeric(sum(total_riders_month, na.rm = TRUE)),

    monthly_trip_based_otp = as.numeric(mean(monthly_trip_otp, na.rm = TRUE)),

    passenger_weighted_monthly_otp = as.numeric(
      sum(total_riders_month * monthly_trip_otp, na.rm = TRUE) /
        sum(total_riders_month, na.rm = TRUE)
    ),

    avg_trip_riders = as.numeric(mean(total_riders_month, na.rm = TRUE)),
    avg_match_share = as.numeric(mean(dominant_share, na.rm = TRUE))
  ),
  by = Route
]

route_monthly_summary[, otp_gap := passenger_weighted_monthly_otp - monthly_trip_based_otp]
route_monthly_summary[, monthly_trip_based_otp_pct := 100 * monthly_trip_based_otp]
route_monthly_summary[, passenger_weighted_monthly_otp_pct := 100 * passenger_weighted_monthly_otp]
route_monthly_summary[, otp_gap_pct_points := 100 * otp_gap]

setorder(route_monthly_summary, Route)

# Save summary table
monthly_visual_summary_file <- file.path(
  output_dir,
  paste0("route_monthly_bridge_summary_", month_tag, "_", gtfs_tag, ".csv")
)

fwrite(route_monthly_summary, file = monthly_visual_summary_file)

# Scatter plot
monthly_scatter_file <- file.path(
  output_dir,
  paste0("monthly_bridge_otp_vs_passenger_weighted_otp_", month_tag, "_", gtfs_tag, ".png")
)

p_monthly <- ggplot(
  route_monthly_summary,
  aes(
    x = monthly_trip_based_otp,
    y = passenger_weighted_monthly_otp,
    size = total_riders_month,
    label = Route
  )
) +
  geom_point(alpha = 0.75) +
  geom_text(nudge_y = 0.01, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Monthly Trip-based OTP vs Passenger-weighted Monthly OTP",
    x = "Monthly Trip-based OTP",
    y = "Passenger-weighted Monthly OTP",
    size = "Total Monthly Riders"
  ) +
  theme_minimal()

ggsave(filename = monthly_scatter_file, plot = p_monthly, width = 8, height = 6, dpi = 300)

cat("Saved monthly bridge summary file:\n", monthly_visual_summary_file, "\n")
cat("Saved monthly scatter plot:\n", monthly_scatter_file, "\n")

print(route_monthly_summary[1:10])

# Optional: show routes with biggest positive/negative gap
cat("\nRoutes where passenger-weighted OTP is LOWER than trip-based OTP:\n")
print(route_monthly_summary[order(otp_gap)][1:10, .(
  Route, n_trip_patterns, total_riders_month,
  monthly_trip_based_otp_pct,
  passenger_weighted_monthly_otp_pct,
  otp_gap_pct_points
)])

cat("\nRoutes where passenger-weighted OTP is HIGHER than trip-based OTP:\n")
print(route_monthly_summary[order(-otp_gap)][1:10, .(
  Route, n_trip_patterns, total_riders_month,
  monthly_trip_based_otp_pct,
  passenger_weighted_monthly_otp_pct,
  otp_gap_pct_points
)])

# Cleanup
rm(
  monthly_bridge, sys_for_monthly, system_monthly_otp,
  monthly_bridge_use, monthly_trip_join,
  route_monthly_summary, p_monthly
)
gc()